---
title: "Using the usheR github easy updater"
author: "Pawel Orzechowski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the usheR github easy updater}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(usheR)
```

# Introduction - Simplified Classnotes Updater (no-drama github pull-and-merge):

This part of the package allows students to 'update/pull' most recent notes from teacher's public course git repository without using git commands in terminal or RStudio. This is especially useful on the beginner courses, and in the first weeks where you want to start using R code quickly, prior to teaching students abotu git.

# How To use it:

## As a Teacher:

-   This assumes you have all your class notes for students in a github repo, which includes the RStudio project
-   Create the student-facing file (instructions notes below, call it `RUN_TO_GET_RECENT_NOTES.Rmd`) from this package in the top folder of your teaching-notes repo.
-   That file installs UsheR package and runs a function from it. That function performs a git commit-merge-pull flow behind the scenes (so that students does not need to). Any conflicts are dealt with gracefully (by keeping both student and teacher copy, with a timestamp).

## As a Students:

-   Once, on the beginning of the course, students will need to clone your teaching repo using RStudio menu interface (using `File > New Project > Version Control > Git` and use your Repository Url).
-   Whenever students want to have the most recent version of the class notes (e.g. every week) they need to open RStudio and run the code block in file `RUN_TO_GET_RECENT_NOTES.Rmd`.
-   This should work both on local RStudio, PositCloud and Noteable.

# How to create a student facing file:

Copy-paste below code into a file and save it at the top level of your public repo with notes. Call the file something like `RUN_TO_GET_RECENT_NOTES.Rmd`.

Whevener students run that file:

-   new files from github will be added (pulled) into their project directory
-   without overwriting their notes
-   without the need for student to log into github (assuming repo is public)

Put ALL of below text in a file `RUN_TO_GET_RECENT_NOTES.Rmd`:

------------------------------------------------------------------------

<!-- start of file RUN_TO_GET_RECENT_NOTES.Rmd-->

Run below code block to get the most recent notes from github.

To run a grey code block below, click anywhere inside it and then click the green play button in top left corner and click.

```{r echo = T, results = 'hide'}
install.packages("devtools")
devtools::install_github('ddi-talent/usheR')

library(usheR)
update_from_github()
```

Once you run it, your R studio will install some packages and bring the most recent course notes from github to your R project.

------------------------------------------------------------------------

<!-- end of file RUN_TO_GET_RECENT_NOTES.Rmd-->
